---
title: "Visualizing more than 3 dimensions"
author: "Niklas Schandry"
date: "May / June 2020"
output:
  github_document: default
---

# Motivation

I am teaching a course on metagenomic analysis, with the core audience being molecular biologists. My goal is to explain data analysis that are commonly use for these multivariate datasets, and how to perform these using R. I am not intending to teach math, or pure stats, but data-science. Part of this is finding low-dimensional representations of a high-dimensional space, basically ordination. 
 
In my experience, two dimensional spaces are easy to grasp, and can be visualized easily as an xy-plot. Some people are able to visualize a three dimensional space, but this can already be challenging for many datapoints. 

Literally everyone (myself included) is highly confused when trying to imagine a four-dimensional space, using axis that point, well, into three dimensional space. It is not possible using another spacial dimension, because we cannot imagine more than three orthogonal spacial dimensions, and trying to do this is our first instinct, as we are already thinking in a "space-space" (? i am no mathematician)

However, anyone who does any type of analysis does in fact regularly project additional dimensions onto a two-dimensional space, for example when we color dots by a categorial or continous variable. We can use the exact same principle, to add a fourth dimension to spheres in a three-dimensional space. 

To explain, we generate a matrix with independent, and less independent dimensions, this matrix is centered. 

```{r}
library(tidyverse)
library(magrittr)
library(patchwork)
```


```{r}
set.seed(238954)
Dim1 = rnorm(10,0,25)
Dim2 = rnorm(10,0,9)
Dim3 = rnorm(10,0,15)
Dim4 = Dim3 * (- 2.4) + rnorm(10,0,7)
cent_mat_2 <- cbind(Dim1,Dim2,Dim3,Dim4) %>% as.data.frame() %>% round(1)
```

```{r}
cent_mat_2
```

We see that Dim1 has the largest variance of the first three (720.4), followed by Dim3 (393.7). However, Dim4 widely exceeds the variance of Dim1 by a factor of ~3.1 (see above)

```{r}
cent_mat_2 %>% var %>% round(1)
```
Initially, we only focus on the first three dimensions.


```{r}
p1_2 <- cent_mat_2 %>% 
  ggplot() +
  geom_point(aes(x=Dim1,y=Dim2))

p1_3 <- cent_mat_2 %>% 
  ggplot() +
  geom_point(aes(x=Dim1,y=Dim3))

p2_3  <- cent_mat_2 %>% 
  ggplot() +
  geom_point(aes(x=Dim2,y=Dim3))

(p1_2 + p1_3 + p2_3)  & theme_minimal()
```

This repository contains a blender file (in /blender) in which i have reproduced the three dimensional arrangement.
This scene (in blender) also contains one very strong light, and a camera, both are centered on 0,0,0. The spheres completely absorb light, and therefore appear as dots in the camera view.
Both, light and camera are fixed in their distance to 0,0,0. There is an "axis" element ("Empty" in the right panel, displayed as a black cross) which is at 0,0,0. Both the camera and the light follow the rotation of this element. This allows us to freely rotate around the center in three dimensional space, observing the "projection" in the camera panel on the lower right.

!(imgs/Blender_scene_overview.PNG)

If we try to mimick a principal component analysis, we rotate (click on the "empty" element and press R and then move around in the panels showing the coordinate system, pressing R twice allows for a more "free" rotation, which is cool and confusing) until we maximize distance along the x axis of this panel, and then "fix" this and rotate to maximize the distance along the other axis. It is not hard to find a projection that approximates the output of a PCA pretty well, once we get a feel for how rotation works in blender. This is me playing a bit in blender:

!(imgs/First_projection.PNG)

This is the PCA result:

```{r}
pca_obj_1 <- cent_mat_2[,1:3] %>% pcaMethods::pca(center = F) 
pca_obj_1@scores %>%
  as.data.frame() %>%
  ggplot(aes(x = PC1, y =  PC2)) + 
  geom_point() +
  theme_minimal()
```
Now to understand how we can project a fourth dimension, that is orthogonal to the other three, we simply map dimension 4 to a color vector. I have tried to explain this using the plots below, initially we map dim1 and dim4 in space, then we color the dots by their dim4 value, and then, we remove the axis that was used for dim4, retaining information about both dimensions in a one-dimensional display.


```{r dim1_2}
p1_4 <- ggplot(cent_mat_2, aes(x = Dim1, y = Dim4)) +
  geom_point() +
  theme_minimal()
```

```{r}
p1_4_color_no_y <- ggplot(cent_mat_2) + 
  geom_point(aes(x = Dim1, y = rep(0,10), color = Dim4)) +
  theme_minimal() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text.y = element_blank()) +
  labs(y = element_blank()) +
  scale_color_viridis_c() 
```

```{r}

```

```{r dim1_4 colored}
p1_4_color <- cent_mat_2 %>%
  ggplot(aes(x = Dim1, y = Dim4)) +
  geom_point(aes(color = Dim4)) +
  theme_minimal() +
  scale_color_viridis_c() 
```


```{r}
p1_4
```

```{r}
p1_4_color
```


```{r}
p1_4_color_no_y
```


I extract these colors.

```{r}
plotdat <- ggplot_build(p1_4_color_no_y)
colors <- plotdat$data[[1]] %$% colour
```

```{r}
colors
```

In the blender, you will notice that there is a "collection" called sample colors 

!(imgs/blender_sample_color_collection.PNG)

This is the same samples, but colored according to their value, you can display this by clicking on the eye-icon to make them visible.

We can now project their position in the fourth dimension as well. We also know (see above) that dimension four explains much more variance. We now try to project dimension four (from yellow to purple) along the X axis in the camera view. Then, we rotate to maximize the distance along the other axis. Since we take into account variance of the color vector explicitly (and that of the other vectors implicitly since we "see" the variance as the distance between points) we are actually able to intuitively maximize these, the hardest part is gaining intuition for how the blender rotation works. This is my projection:

!(imgs/Second_Projection.PNG)

Again, we can compare this to the PCA result on these four columns.

```{r}
pca_obj_1 <- cent_mat_2 %>% pcaMethods::pca(center = F) 
pca_obj_1@scores %>%
  as.data.frame() %>%
  ggplot(aes(x = PC1, y =  PC2)) + 
  geom_point(color=colors)+
  theme_minimal()
```

Once we understand this, we can actually imagine even more dimension, using common mappings e.g. shape, or transparency to add additional dimensions beyond the third.