---
title: "PCA_Expl"
author: "Niklas Schandry"
date: "May / June 2020"
output:
  pdf_document:
    toc: yes
  word_document:
    toc: yes
  html_document:
    fig_caption: yes
    fig_height: 9
    fig_width: 16
    highlight: haddock
    theme: cosmo
    toc: yes
    toc_float: yes
---

# Motivation

I am teaching a course on metagenomic analysis, with the core audience being molecular biologists. My goal is to explain data analysis that are commonly use for these multivariate datasets, and how to perform these using R. I am not intending to teach math, or pure stats, but data-science. Part of this is finding low-dimensional representations of a high-dimensional space, basically ordination. 
 
In my experience, two dimensional spaces are easy to grasp, and can be visualized easily as an xy-plot. Some people are able to visualize a three dimensional space, but this can already be challenging for many datapoints. 

Literally everyone (myself included) is highly confused when trying to imagine a four-dimensional space, using axis that point, well, into three dimensional space. It is not possible using another spacial dimension, because we cannot imagine more than three orthogonal spacial dimensions, and trying to do this is our first instinct, as we are already thinking in a "space-space" (? i am no mathematician)

However, anyone who does any type of analysis does in fact regularly project additional dimensions onto a two-dimensional space, for example when we color dots by a categorial or continous variable. We can use the exact same principle, to add a fourth dimension to spheres in a three-dimensional space. 

```{r}
library(tidyverse)
library(magrittr)
library(patchwork)
```

# Data in

```{r}
dist_mat <- readRDS("dat/distmat.rds")

cent_mat <- readRDS("dat/centmat.rds")
```

```{r}
cent_mat %>% cov()
```

```{r}
p1 <- ggplot(dist_mat, aes(x = Sample1, y = Sample2)) +
  geom_point()
p2 <- ggplot(dist_mat, aes(x = Sample1, y = Sample3)) +
  geom_point()
p3 <- ggplot(dist_mat, aes(x = Sample1, y = Sample4)) +
  geom_point()
p4 <- ggplot(dist_mat, aes(x = Sample3, y = Sample4)) +
  geom_point()
p1+p2+p3+p4
```
```{r}
p1 <- ggplot(dist_mat, aes(x = Sample1, y = rep(0,10))) +
  geom_point() +theme_minimal() +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text.y = element_blank()) +
  labs(y = element_blank())
  

p2 <- ggplot(dist_mat, aes(x = rep(0,10), y = Sample2 )) +
  geom_point() +theme_minimal() +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text.x = element_blank()) +
  labs(x = element_blank())
p3 <- ggplot(dist_mat, aes(x = Sample1, y = Sample2)) +
  geom_point() +
  theme_minimal()
p4 <- ggplot(dist_mat, aes(x = Sample3, y = Sample4)) +
  geom_point()
(p1+p2) / p3
```


A centered version of this matrix, where the axis are centered to their mean

```{r}
dist_mat[dist_mat == 0] <- NA
cent_mat <- dist_mat %>% mutate(Sample1 = Sample1 - mean(Sample1, na.rm = T) %>% round(1),
                   Sample2 = Sample2 - mean(Sample2, na.rm = T) %>% round(1),
                   Sample3 = Sample3 - mean(Sample3, na.rm = T) %>% round(1),
                   Sample4 = Sample4 - mean(Sample4, na.rm = T) %>% round(1))
cent_mat[1,1] <- -mean(dist_mat[,1], na.rm = T)
cent_mat[2,2] <- -mean(dist_mat[,2], na.rm = T)
cent_mat[3,3] <- -mean(dist_mat[,3], na.rm = T)
cent_mat[4,4] <- -mean(dist_mat[,4], na.rm = T)

cent_mat <- cent_mat %>% round(1)
```


## Using a new matrix

The following examples are done using a different matrix. The distance matrix is correlated, in all dimensions. This hides some problems. This is also available as blender file, with xyz corresponding to 1,2,3 and color corresponding to 4. 

```{r}
set.seed(238954)
Dim1 = rnorm(10,0,25)
Dim2 = rnorm(10,0,9)
Dim3 = rnorm(10,0,15)
Dim4 = Dim3 * (- 2.4) + rnorm(10,0,7)
cent_mat_2 <- cbind(Dim1,Dim2,Dim3,Dim4) %>% as.data.frame() %>% round(1)
```

```{r}
cent_mat_2
```


```{r}
p1_2 <- cent_mat_2 %>% 
  ggplot() +
  geom_point(aes(x=Dim1,y=Dim2))

p1_3 <- cent_mat_2 %>% 
  ggplot() +
  geom_point(aes(x=Dim1,y=Dim3))

p1_4 <- cent_mat_2 %>% 
  ggplot() +
  geom_point(aes(x=Dim1,y=Dim4))

p1_2 + p1_3 + p1_4
```

```{r}
p2_3 <- cent_mat_2 %>% 
  ggplot() +
  geom_point(aes(x=Dim2,y=Dim3))

p2_4 <- cent_mat_2 %>% 
  ggplot() +
  geom_point(aes(x=Dim2,y=Dim4))

p3_4  <- cent_mat_2 %>% 
  ggplot() +
  geom_point(aes(x=Dim3,y=Dim4))

(p1_2 + p1_3 + p2_3)  & theme_minimal()
```



```{r}
p_3_4_c <- ggplot(cent_mat_2, aes(x = Dim1, y = Dim4)) + geom_point() + theme_minimal()

p_3_4_w_color_c <- ggplot(cent_mat_2, aes(x = Dim1, y = Dim4)) + geom_point(aes(color = Dim4)) + theme_minimal() + scale_colour_viridis_c()

p_3_4_w_color_no_y_c <- ggplot(cent_mat_2, aes(x = Dim1, y = rep(0,10))) + 
  geom_point(aes(color = Dim4)) +
  theme_minimal() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text.y = element_blank()) +
  scale_colour_viridis_c() + labs(y = element_blank())

(p_3_4_c + p_3_4_w_color_c) / p_3_4_w_color_no_y_c
```

```{r}
plotdat <- ggplot_build(p_3_4_w_color_no_y_c)
colors <- plotdat$data[[1]] %$% colour
```

```{r}
colors
```

```{r}
pca_obj_1 <- cent_mat_2[,1:3] %>% pcaMethods::pca(center = F) 
pca_obj_1@scores %>%
  as.data.frame() %>%
  ggplot(aes(x = PC1, y =  PC2)) + 
  geom_point() +
  theme_minimal()
```
```{r}
pca_obj_1 <- cent_mat_2 %>% pcaMethods::pca(center = F) 
pca_obj_1@scores %>%
  as.data.frame() %>%
  ggplot(aes(x = PC1, y =  PC2)) + 
  geom_point(color=colors)+
  theme_minimal()
```

